Factoring with $n+2$ clean qubits and $n-1$ dirty qubits

### 2017-06-01 doc/paper.tex
License headers
Update document style - Remove emphasis from figure captions - Use autoref for figure/section references - Use quantumarticle style - Add affiliation / email - Blue links

\documentclass[twocolumn]{article}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{hyperref}
\title{Factoring with $n+2$ clean qubits and $n-1$ dirty qubits}
\author{Craig Gidney}

@@

\begin{abstract}
\em
We present reversible classical circuits for performing various arithmetic operations aided by dirty ancillae in an unknown state that must be preserved.
We improve the number of clean qubits needed to factor an $n$-bit number with Shor's algorithm \cite{Shor1999} from $2n+2$ \cite{takahashi2006, haner2016} to $n+2$, and the total number of qubits needed from $2n+2$ to $2n+1$, without increasing the asymptotic size or depth of the circuit.
\end{abstract}

@@

Our paper is structured as follows.
Section \ref{sec:introduction} introduces and describes the conventions our circuit constructions and circuit diagrams will follow.
In section \ref{sec:construct} we describe all the circuit constructions we use to reduce the period finding step of Shor's algorithm into constant-sized gates, while tracking the number of required dirty ancillae.
Then, in section \ref{sec:costs}, we discuss the novelty and comparative costs of the presented circuit constructions.
Finally, section \ref{sec:conclusion} concludes with a discussion of possible future improvements.

@@

The ancillae counts shown in diagrams and discussed in the text are not optimal; they are entirely based on what the constructions we discuss in this paper achieve.
Nearly all the circuit constructions we present are classical reversible circuits (i.e. they do not use any quantum operations), so tricks such as the ancilla bootstrap shown in figure \ref{fig:bootstrap-ancilla} could cut the ancilla count in several places.
However, none of these improvements decrease the ancillae required at any crucial bottleneck, and so they do not reduce the number of qubits required by our overall construction of Shor's algorithm.

To avoid ambiguity, circuit diagrams will divide multi-register operations into separate parts.
For example, for an addition operation that adds a register $x$ into a register $y$ (i.e. performs $|x\rangle |y\rangle \rightarrow |x\rangle |y + x \pmod{2^n} \rangle$), we will place a light-gray box with the text ``Input A" over the wires corresponding to $x$ and a white box with the text ``+A" over the wires corresponding to $y$.
The ``A" symbol refers to the value of the input register, in the computational basis, at the time of the operation.
Diagrams will often use a sequence of operations each with their own ``Input A" box specific to that operation.
This re-use of ``A" does not indicate any relation between those input values.
The use of ``A" is merely a convention for indicating how the input-indicating box and the effect-indicating box are supposed to combine to form a single operation.

@@

Our reduction, from the period finding step of Shor's algorithm down to constant-sized gates, uses many circuit constructions.
Figure \ref{fig:dependencies} shows an overview of the various operations we will use, and the path through them that we will take.

\begin{figure}
  \centering
  \includegraphics[height=17.3cm]{assets/dependencies.png}
  \caption{\em
    Graph of the transitive reduction of dependencies between constructions in our paper.
    Edge labels indicate which constructions use a dependency more than a constant number of times.
  }
  \label{fig:dependencies}
\end{figure}

@@

\begin{figure}
  \centering
  \makebox[\linewidth]{
    \includegraphics[width=\linewidth]{assets/shor-period-finding.png}
  }
  \caption{\em
	High-level definition of the period finding circuit \cite{Shor1999} used by Shor's algorithm.
	$R$ is the modulus and the number to be factored, $B$ is a randomly chosen base, $n$ is the number of bits needed to store $R$, and $p \in \Theta(n)$ controls the precision of the phase estimation step.
    The triangles indicate how many ancillae are needed ``behind the scenes", by our constructions, to perform an operation.
	Recovering the period requires classical post-processing of the sampled output.
  }
  \label{fig:period-finding}
\end{figure}

A high-level view of the circuit for period finding applied to modular exponentiation, the core quantum subroutine of Shor's quantum factoring algorithm \cite{Shor1999}, is shown in figure \ref{fig:period-finding}.

@@

Because period-finding measures all qubits immediately after performing a QFT, most of the transformed qubits can be measured earlier than shown.
In fact, each qubit can be measured so early that the next qubit needed for the QFT does not even need to be initialized yet!
Only one of the qubits in the phase-estimation register needs to be present at a time, and so the phase register can be reduced to a single repeatedly-used qubit \cite{zalka1998, mosca1999, parker2000, beauregard2003}.
Figure \ref{fig:period-finding-solo-phase-qubit} shows a period-finding circuit with this property.
It uses (a) controlled modular multiplication, (b) measurement, (c) X-axis rotations classically parametrized by previous measurements, and (d) qubit resets.
The only non-trivial operation is (a), the controlled modular multiplication of an $n$-qubit register.

@@

The bimultiplications we perform throughout the algorithm do change the value of the ancilla register, but after the usual end of the circuit we can undo the damage.
The key insight is that the work register is initialized to $|1\rangle$ and gets multiplied by constants inverse to the ones trashing the dirty ancilla register.
Instead of discarding the work register after performing phase estimation, as is normally done, we have a use for it.
We measure the work register to recover the fixup factor needed to restore the dirty register's original value.
Figure \ref{fig:period-finding-solo-phase-qubit-explicit-dirty-register} shows this construction.

@@

\begin{figure*}
  \centering
  \makebox[\linewidth]{
    \includegraphics[width=\linewidth]{assets/shor-period-finding-solo-phase-qubit.png}
  }
  \caption{\em
	Period finding with a single phase-estimation qubit \cite{beauregard2003}.
	The small oplus' ({\tiny $\oplus$}) are ``X-axis controls".
	An X-axis control is equivalent to a normal control, but with a Hadamard gate applied before and after.
	It conditions on the state $\frac{1}{\sqrt 2}|0\rangle - \frac{1}{\sqrt 2}|1\rangle$ instead of on the state $|1\rangle$.
  }
  \label{fig:period-finding-solo-phase-qubit}
\end{figure*}

\begin{figure*}
  \centering
  \makebox[\linewidth]{
    \includegraphics[width=\linewidth]{assets/shor-period-finding-solo-phase-qubit-double-register.png}
  }
  \caption{\em
	Period finding with a single phase-estimation qubit and paired inverse multiplications (``bimultiplications").
	Uses $O(n^3 \lg n)$ gates, $O(n^3)$ depth, and no additional qubits beyond those shown.
  }
  \label{fig:period-finding-solo-phase-qubit-explicit-dirty-register}
\end{figure*}

@@

As in \cite{beauregard2003}, we perform controlled modular multiplication with a second register and modular scaled-addition operations.
However, to allow the second register to be dirty, we use an additional scale-add as well as a negation operation.
See figure \ref{fig:controlled-modular-multiply} for the circuit diagram.

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-modular-multiply.png}
  \caption{\em
    Controlled modular bimultiplication, with a constant multiplier $K$, using three modular scaled-additions and a swap.
    $K$ must have a multiplicative inverse modulo $R$.
    Uses no ancillae, $O(c + n^2 \lg n)$ gates, and $O(c + n^2)$ depth where $c$ is the number of controls and $n$ is the register size.
    The negation operation requires 2 dirty ancillae, but notice there are free wires in the other register that can be borrowed, so the overall construction requires no additional dirty ancilla (for non-trivial $n$).
  }
  \label{fig:controlled-modular-multiply}
\end{figure}

@@

To perform modular scale-add operations, we use a shift-and-add approach similar to \cite{beauregard2003}.
We right-shift (i.e. divide by $2 {\pmod R}$) the target register $n-1$ times, then begin iteratively left-shifting (i.e. multiplying by $2 {\pmod R}$) and adding $K$ into the target.
We condition the first modular addition on the most significant bit of the input, the second addition on the next most significant bit, and so forth.
(The more significant bits go first because their effects must be hit by more left-shifts.)
See figure \ref{fig:controlled-modular-scale-accumulate}.

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-modular-multiply-accumulate.png}
  \caption{\em
    Reducing a controlled modular scale-add, with a constant multiplier $K$, into the modular equivalent of shift-and-add.
    Requires the modulus $R$ to be odd.
    Uses no ancillae, $O(c + n^2 \lg n)$ gates, and $O(c + n^2)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:controlled-modular-scale-accumulate}
\end{figure}

@@

Because it's expensive to perform an offset that affects values above $R/2$ while leaving values below $R/2$ alone, the circuit shown in figure \ref{fig:modular-double} uses a different alignment strategy.
It starts by offset-ing the whole range down by $\lceil R/2 \rceil$, which wraps the valid-bottom-half of the input around to the top of the register's range.
This offset also aligns what was originally the valid-top-half against the bottom of the register range.
Then the circuit uses a controlled offset to move just the top half of the register (including the invalid range above $R$) up by $\lceil R/2 \rceil$, re-wrapping what was originally the valid-bottom-half past the top and around to the middle, aligning it with the MSB boundary.
The circuit then toggles the MSB, fixing the fact that, although the valid-top-half is aligned with 0 and the valid-bottom-half is aligned with $2^{n-1}$, that's exactly the reverse of what we want.
With the desired alignment achieved, the circuit interleaves the two halves with a left-rotate.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-modular-double.png}
  \caption{\em
    Controlled modular doubling.
    $R$ must be odd.
    Uses 1 dirty ancilla, $O(c + n \lg n)$ gates, and $O(c + n)$ depth, where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:modular-double}
\end{figure}

@@

See figures \ref{fig:controlled-pivot-flip} and \ref{fig:controlled-const-pivot-flip} for the circuit diagrams.
For states that are less than the pivot, the comparison against the pivot keeps toggling the ancilla and exactly one of the controlled bi-flips will fire.
For states equal to or larger than the pivot, the ancilla does not get toggled and so the bi-flips undo each other.
Therefore only the range below the pivot is flipped.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-pivot-flip.png}
  \caption{\em
    Controlled pivot-flip circuit with an enregistered pivot.
    The target register can be larger than the input register, but not smaller.
    Uses $2$ dirty ancillae, $O(c + m \lg m)$ gates, and $O(c + m)$ depth where $c$ is the number of controls, $n$ is the size of the input register, $e$ is the extra size of the target register, and $m=n+e$ is the size of the target register.
  }
  \label{fig:controlled-pivot-flip}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-const-pivot-flip.png}
  \caption{\em
    Controlled pivot-flip circuit with a constant pivot.
    Uses $2$ dirty ancillae, $O(c + n \lg n)$ gates, and $O(c + n)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:controlled-const-pivot-flip}
\end{figure}

@@

As shown in figure \ref{fig:mod-add-from-pivot-flip-bars}, a modular addition can be implemented by three pivot-flips.
To add $K$ into a register modulo $R$, perform pivot-flips with the pivot at $R-K$, then $R$, then $K$.
See figure \ref{fig:controlled-modular-add} for the circuit.
(An optimization we don't show is that, because the values above $R$ don't matter, the pivot-flip at $R$ can be replaced by a bi-flip.)

Interestingly, controlled modular addition can borrow its own controls as dirty bits.
For modular offset (i.e. adding a compile-time constant into a register), the two dirty bits are required whether or not the operation is controlled.
See figure \ref{fig:controlled-modular-offset}.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/mod-add-from-pivot-flip-bars.png}
  \caption{\em
     Modular addition of $A \pmod{R}$ can be done with three pivot flips.
     One at $R-A$, then one at $R$, then one at $A$.
     Requires $A \leq R$.
   }
  \label{fig:mod-add-from-pivot-flip-bars}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-modular-addition.png}
  \caption{\em
    Controlled modular addition construction based on pivot-flips.
    The arithmetic being performed on the input register temporarily transitions it from storing $x$ to $R-x$ for the first pivot flip.
    Because $R$ is a compile-time constant, $R+1$ is also a compile-time constant and the $+R+1$ operation is a normal offset operation.
    All arithmetic is two's complement, i.e. modulo $2^n$.
    Uses $2-c$ dirty ancillae, $O(c + n \lg n)$ gates, and $O(c + n)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:controlled-modular-add}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-modular-offset.png}
  \caption{\em
    Controlled modular offset construction based on pivot-flips.
    Uses $2$ dirty ancillae, $O(c + n \lg n)$ gates, and $O(c + n)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:controlled-modular-offset}
\end{figure}

@@

To negate a number mod $R$, we need to reverse the order of the states $|1\rangle$ to $|R-1\rangle$.
We do so by temporarily moving $|0\rangle$ out the way with a decrement, pivot-flipping at $R-1$, then undoing the decrement.
See figure \ref{fig:negate-mod}.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/negate-mod.png}
  \caption{\em
    Controlled modular negation.
    Uses $2$ dirty ancillae, $O(c + n \lg n)$ gates, and $O(c + n)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:negate-mod}
\end{figure}

@@

Comparison operations toggle a target bit based on the relationship between two input registers.
We implement comparisons as in \cite{takahashi2005}, using an addition followed by a slightly smaller subtraction that clears all changes except the overflow signal into the target bit.
See figure \ref{fig:compare}.

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/comparison-less-const.png}
  \caption{\em
    Controlled comparison.
    Uses $2$ dirty ancillae, $O(c + n \lg n)$ gates, and $O(c + n)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:compare}
\end{figure}

@@

\cite{takahashi2005} provides a reversible adder circuit that uses $O(n)$ gates, $O(n)$ depth, and no ancillae.
We show an equivalent circuit in figure \ref{fig:inlineadder}.

Because the construction in figure \ref{fig:inlineadder} uses the input register as workspace, it doesn't work when the input is a compile-time constant or when the target register is larger than the input register.
(When the target register is smaller, we simply ignore the high bits of the input due to overflow wraparound.)
This is a problem for us, because some of our constructions add constants into register (e.g. modular scale-add), and some others add inputs into larger target registers (e.g. comparison).

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/inline-adder.png}
  \caption{\em Adder for input and target registers of the same size.
  Requires no ancillae, and uses $O(N)$ gates and depth.
  Based on \cite{van2004, takahashi2005}.}
  \label{fig:inlineadder}
\end{figure}

When the value to be added into a register is a compile-time constant (i.e. when applying an offset gate), we use the offset construction from \cite{haner2016}.
Their offset circuit, shown in figure \ref{fig:offset}, uses $O(n \lg n)$ gates, $O(n)$ depth, and one dirty ancilla.

When the target register is larger than the input register, the addition circuit in figure \ref{fig:inlineadder} can be modified without increasing the asymptotic cost.
We do this by removing the surrounding CNOTs and adding three increment/decrement operations to the circuit, as follows.
(Note that, to avoid cyclic dependencies, the increment and decrement constructions described in the next subsection will only use the same-register-size adder.)

First, because there are many target bits not reached by the time we have swept through the entire input register, carry signals aren't reaching the high bit of the target register.
We fix this by replacing the innermost CNOT, the carry-propagating CNOT, with a controlled increment.
Second and third, because we're using the MSB of the input register as the carry signal, it causes an increment at the LSB instead of at the correct position.
We undo the LSB increment with a controlled decrement, and use a controlled increment to apply the MSB's effect where it should have actually gone.
See the circuit diagram in figure \ref{fig:inline-adder-into-large}.

@@

In the case of addition, we have the property $\lnot (\lnot x + K) = \lnot (-x - 1 + K) = -(-x - 1 + K) - 1 = x-K$.
So a valid inverting operation $H$ for addition is a NOT gate applied to every target wire; $H=X^{\otimes n}$.
If we then consider $G=\text{S}_{\text{-} K}$, where $S$ is an offset-by-subscript operation, we find that $[G, H] = [\text{S}_{-K}, X^{\otimes n}] = \text{S}_{2K}$.
We can control this constructed $\text{S}_{2K}$ operation by adding controls only to the multi-not operation $H$, which is cheap to do.
To add $K$ instead of $2K$, we temporarily prepend a dirty LSB onto the target register.
We show this construction in figure \ref{fig:controlled-addition}.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/offset.png}
  \caption{\em
      Offset circuit from \cite{haner2016}.
      $e$ is either 0 or 1.
      Uses $1$ dirty ancilla, $O(n \lg n)$ gates, and $O(n)$ depth (by overlapping the recursive cases).
  }
  \label{fig:offset}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/inline-adder-into-large.png}
  \caption{\em
      Adder with target larger than source, using no ancillae.
      Uses $O(n)$ gates and depth.
      The increment and decrement gates all have at least one free wire to borrow as a dirty ancilla.
  }
  \label{fig:inline-adder-into-large}
\end{figure}

>>>>

\documentclass[twocolumn,longbibliography]{quantumarticle}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[pdfpagelabels,pdftex,bookmarks,breaklinks]{hyperref}
\hypersetup{colorlinks,citecolor=blue,urlcolor=blue,linkcolor=blue}

\title{Factoring with $n+2$ clean qubits and $n-1$ dirty qubits}
\author{Craig Gidney}
\affiliation{Google, Santa Barbara, CA 93117, USA}
\email{craiggidney@google.com}

\def\sectionautorefname{Section}

@@

\begin{abstract}
We present reversible classical circuits for performing various arithmetic operations aided by dirty ancillae in an unknown state that must be preserved.
We improve the number of clean qubits needed to factor an $n$-bit number with Shor's algorithm \cite{Shor1999} from $2n+2$ \cite{takahashi2006, haner2016} to $n+2$, and the total number of qubits needed from $2n+2$ to $2n+1$, without increasing the asymptotic size or depth of the circuit.
\end{abstract}

@@

Our paper is structured as follows.
\autoref{sec:introduction} introduces and describes the conventions our circuit constructions and circuit diagrams will follow.
In \autoref{sec:construct} we describe all the circuit constructions we use to reduce the period finding step of Shor's algorithm into constant-sized gates, while tracking the number of required dirty ancillae.
Then, in \autoref{sec:costs}, we discuss the novelty and comparative costs of the presented circuit constructions.
Finally, \autoref{sec:conclusion} concludes with a discussion of possible future improvements.

@@

The ancillae counts shown in diagrams and discussed in the text are not optimal; they are entirely based on what the constructions we discuss in this paper achieve.
Nearly all the circuit constructions we present are classical reversible circuits (i.e. they do not use any quantum operations), so tricks such as the ancilla bootstrap shown in \autoref{fig:bootstrap-ancilla} could cut the ancilla count in several places.
However, none of these improvements decrease the ancillae required at any crucial bottleneck, and so they do not reduce the number of qubits required by our overall construction of Shor's algorithm.

To avoid ambiguity, circuit diagrams will divide multi-register operations into separate parts.
For example, for an addition operation that adds a register $x$ into a register $y$ (i.e. performs $|x\rangle |y\rangle \rightarrow |x\rangle |y + x \pmod{2^n} \rangle$), we will place a light-gray box with the text ``Input A" over the wires corresponding to $x$ and a white box with the text ``+A" over the wires corresponding to $y$.
The ``A" symbol refers to the value of the input register, in the computational basis, at the time of the operation.
Diagrams will often use a sequence of operations each with their own ``Input A" box specific to that operation.
This reuse of ``A" does not indicate any relation between those input values.
The use of ``A" is merely a convention for indicating how the input-indicating box and the effect-indicating box are supposed to combine to form a single operation.

@@

Our reduction, from the period finding step of Shor's algorithm down to constant-sized gates, uses many circuit constructions.
\autoref{fig:dependencies} shows an overview of the various operations we will use, and the path through them that we will take.

\begin{figure}
  \centering
  \includegraphics[height=17.3cm]{assets/dependencies.png}
  \caption{
    Graph of the transitive reduction of dependencies between constructions in our paper.
    Edge labels indicate which constructions use a dependency more than a constant number of times.
  }
  \label{fig:dependencies}
\end{figure}

@@

\begin{figure}
  \centering
  \makebox[\linewidth]{
    \includegraphics[width=\linewidth]{assets/shor-period-finding.png}
  }
  \caption{
	High-level definition of the period finding circuit \cite{Shor1999} used by Shor's algorithm.
	$R$ is the modulus and the number to be factored, $B$ is a randomly chosen base, $n$ is the number of bits needed to store $R$, and $p \in \Theta(n)$ controls the precision of the phase estimation step.
    The triangles indicate how many ancillae are needed ``behind the scenes", by our constructions, to perform an operation.
	Recovering the period requires classical post-processing of the sampled output.
  }
  \label{fig:period-finding}
\end{figure}

A high-level view of the circuit for period finding applied to modular exponentiation, the core quantum subroutine of Shor's quantum factoring algorithm \cite{Shor1999}, is shown in \autoref{fig:period-finding}.

@@

Because period-finding measures all qubits immediately after performing a QFT, most of the transformed qubits can be measured earlier than shown.
In fact, each qubit can be measured so early that the next qubit needed for the QFT does not even need to be initialized yet!
Only one of the qubits in the phase-estimation register needs to be present at a time, and so the phase register can be reduced to a single repeatedly-used qubit \cite{zalka1998, mosca1999, parker2000, beauregard2003}.
\autoref{fig:period-finding-solo-phase-qubit} shows a period-finding circuit with this property.
It uses (a) controlled modular multiplication, (b) measurement, (c) X-axis rotations classically parametrized by previous measurements, and (d) qubit resets.
The only non-trivial operation is (a), the controlled modular multiplication of an $n$-qubit register.

@@

The bimultiplications we perform throughout the algorithm do change the value of the ancilla register, but after the usual end of the circuit we can undo the damage.
The key insight is that the work register is initialized to $|1\rangle$ and gets multiplied by constants inverse to the ones trashing the dirty ancilla register.
Instead of discarding the work register after performing phase estimation, as is normally done, we have a use for it.
We measure the work register to recover the fixup factor needed to restore the dirty register's original value.
\autoref{fig:period-finding-solo-phase-qubit-explicit-dirty-register} shows this construction.

@@

\begin{figure*}
  \centering
  \makebox[\linewidth]{
    \includegraphics[width=\linewidth]{assets/shor-period-finding-solo-phase-qubit.png}
  }
  \caption{
	Period finding with a single phase-estimation qubit \cite{beauregard2003}.
	The small oplus' ({\tiny $\oplus$}) are ``X-axis controls".
	An X-axis control is equivalent to a normal control, but with a Hadamard gate applied before and after.
	It conditions on the state $\frac{1}{\sqrt 2}|0\rangle - \frac{1}{\sqrt 2}|1\rangle$ instead of on the state $|1\rangle$.
  }
  \label{fig:period-finding-solo-phase-qubit}
\end{figure*}

\begin{figure*}
  \centering
  \makebox[\linewidth]{
    \includegraphics[width=\linewidth]{assets/shor-period-finding-solo-phase-qubit-double-register.png}
  }
  \caption{
	Period finding with a single phase-estimation qubit and paired inverse multiplications (``bimultiplications").
	Uses $O(n^3 \lg n)$ gates, $O(n^3)$ depth, and no additional qubits beyond those shown.
  }
  \label{fig:period-finding-solo-phase-qubit-explicit-dirty-register}
\end{figure*}

@@

As in \cite{beauregard2003}, we perform controlled modular multiplication with a second register and modular scaled-addition operations.
However, to allow the second register to be dirty, we use an additional scale-add as well as a negation operation.
See \autoref{fig:controlled-modular-multiply} for the circuit diagram.

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-modular-multiply.png}
  \caption{
    Controlled modular bimultiplication, with a constant multiplier $K$, using three modular scaled-additions and a swap.
    $K$ must have a multiplicative inverse modulo $R$.
    Uses no ancillae, $O(c + n^2 \lg n)$ gates, and $O(c + n^2)$ depth where $c$ is the number of controls and $n$ is the register size.
    The negation operation requires 2 dirty ancillae, but notice there are free wires in the other register that can be borrowed, so the overall construction requires no additional dirty ancilla (for non-trivial $n$).
  }
  \label{fig:controlled-modular-multiply}
\end{figure}

@@

To perform modular scale-add operations, we use a shift-and-add approach similar to \cite{beauregard2003}.
We right-shift (i.e. divide by $2 {\pmod R}$) the target register $n-1$ times, then begin iteratively left-shifting (i.e. multiplying by $2 {\pmod R}$) and adding $K$ into the target.
We condition the first modular addition on the most significant bit of the input, the second addition on the next most significant bit, and so forth.
(The more significant bits go first because their effects must be hit by more left-shifts.)
See \autoref{fig:controlled-modular-scale-accumulate}.

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-modular-multiply-accumulate.png}
  \caption{
    Reducing a controlled modular scale-add, with a constant multiplier $K$, into the modular equivalent of shift-and-add.
    Requires the modulus $R$ to be odd.
    Uses no ancillae, $O(c + n^2 \lg n)$ gates, and $O(c + n^2)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:controlled-modular-scale-accumulate}
\end{figure}

@@

Because it's expensive to perform an offset that affects values above $R/2$ while leaving values below $R/2$ alone, the circuit shown in \autoref{fig:modular-double} uses a different alignment strategy.
It starts by offset-ing the whole range down by $\lceil R/2 \rceil$, which wraps the valid-bottom-half of the input around to the top of the register's range.
This offset also aligns what was originally the valid-top-half against the bottom of the register range.
Then the circuit uses a controlled offset to move just the top half of the register (including the invalid range above $R$) up by $\lceil R/2 \rceil$, re-wrapping what was originally the valid-bottom-half past the top and around to the middle, aligning it with the MSB boundary.
The circuit then toggles the MSB, fixing the fact that, although the valid-top-half is aligned with 0 and the valid-bottom-half is aligned with $2^{n-1}$, that's exactly the reverse of what we want.
With the desired alignment achieved, the circuit interleaves the two halves with a left-rotate.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-modular-double.png}
  \caption{
    Controlled modular doubling.
    $R$ must be odd.
    Uses 1 dirty ancilla, $O(c + n \lg n)$ gates, and $O(c + n)$ depth, where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:modular-double}
\end{figure}

@@

See \autoref{fig:controlled-pivot-flip} and \autoref{fig:controlled-const-pivot-flip} for the circuit diagrams.
For states that are less than the pivot, the comparison against the pivot keeps toggling the ancilla and exactly one of the controlled bi-flips will fire.
For states equal to or larger than the pivot, the ancilla does not get toggled and so the bi-flips undo each other.
Therefore only the range below the pivot is flipped.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-pivot-flip.png}
  \caption{
    Controlled pivot-flip circuit with an enregistered pivot.
    The target register can be larger than the input register, but not smaller.
    Uses $2$ dirty ancillae, $O(c + m \lg m)$ gates, and $O(c + m)$ depth where $c$ is the number of controls, $n$ is the size of the input register, $e$ is the extra size of the target register, and $m=n+e$ is the size of the target register.
  }
  \label{fig:controlled-pivot-flip}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-const-pivot-flip.png}
  \caption{
    Controlled pivot-flip circuit with a constant pivot.
    Uses $2$ dirty ancillae, $O(c + n \lg n)$ gates, and $O(c + n)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:controlled-const-pivot-flip}
\end{figure}

@@

As shown in \autoref{fig:mod-add-from-pivot-flip-bars}, a modular addition can be implemented by three pivot-flips.
To add $K$ into a register modulo $R$, perform pivot-flips with the pivot at $R-K$, then $R$, then $K$.
See \autoref{fig:controlled-modular-add} for the circuit.
(An optimization we don't show is that, because the values above $R$ don't matter, the pivot-flip at $R$ can be replaced by a bi-flip.)

Interestingly, controlled modular addition can borrow its own controls as dirty bits.
For modular offset (i.e. adding a compile-time constant into a register), the two dirty bits are required whether or not the operation is controlled.
See \autoref{fig:controlled-modular-offset}.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/mod-add-from-pivot-flip-bars.png}
  \caption{
     Modular addition of $A \pmod{R}$ can be done with three pivot flips.
     One at $R-A$, then one at $R$, then one at $A$.
     Requires $A \leq R$.
   }
  \label{fig:mod-add-from-pivot-flip-bars}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-modular-addition.png}
  \caption{
    Controlled modular addition construction based on pivot-flips.
    The arithmetic being performed on the input register temporarily transitions it from storing $x$ to $R-x$ for the first pivot flip.
    Because $R$ is a compile-time constant, $R+1$ is also a compile-time constant and the $+R+1$ operation is a normal offset operation.
    All arithmetic is two's complement, i.e. modulo $2^n$.
    Uses $2-c$ dirty ancillae, $O(c + n \lg n)$ gates, and $O(c + n)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:controlled-modular-add}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/controlled-modular-offset.png}
  \caption{
    Controlled modular offset construction based on pivot-flips.
    Uses $2$ dirty ancillae, $O(c + n \lg n)$ gates, and $O(c + n)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:controlled-modular-offset}
\end{figure}

@@

To negate a number mod $R$, we need to reverse the order of the states $|1\rangle$ to $|R-1\rangle$.
We do so by temporarily moving $|0\rangle$ out the way with a decrement, pivot-flipping at $R-1$, then undoing the decrement.
See \autoref{fig:negate-mod}.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/negate-mod.png}
  \caption{
    Controlled modular negation.
    Uses $2$ dirty ancillae, $O(c + n \lg n)$ gates, and $O(c + n)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:negate-mod}
\end{figure}

@@

Comparison operations toggle a target bit based on the relationship between two input registers.
We implement comparisons as in \cite{takahashi2005}, using an addition followed by a slightly smaller subtraction that clears all changes except the overflow signal into the target bit.
See \autoref{fig:compare}.

@@

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/comparison-less-const.png}
  \caption{
    Controlled comparison.
    Uses $2$ dirty ancillae, $O(c + n \lg n)$ gates, and $O(c + n)$ depth where $c$ is the number of controls and $n$ is the register size.
  }
  \label{fig:compare}
\end{figure}

@@

\cite{takahashi2005} provides a reversible adder circuit that uses $O(n)$ gates, $O(n)$ depth, and no ancillae.
We show an equivalent circuit in \autoref{fig:inlineadder}.

Because the construction in \autoref{fig:inlineadder} uses the input register as workspace, it doesn't work when the input is a compile-time constant or when the target register is larger than the input register.
(When the target register is smaller, we simply ignore the high bits of the input due to overflow wraparound.)
This is a problem for us, because some of our constructions add constants into register (e.g. modular scale-add), and some others add inputs into larger target registers (e.g. comparison).

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/inline-adder.png}
  \caption{ Adder for input and target registers of the same size.
  Requires no ancillae, and uses $O(N)$ gates and depth.
  Based on \cite{van2004, takahashi2005}.}
  \label{fig:inlineadder}
\end{figure}

When the value to be added into a register is a compile-time constant (i.e. when applying an offset gate), we use the offset construction from \cite{haner2016}.
Their offset circuit, shown in \autoref{fig:offset}, uses $O(n \lg n)$ gates, $O(n)$ depth, and one dirty ancilla.

When the target register is larger than the input register, the addition circuit in \autoref{fig:inlineadder} can be modified without increasing the asymptotic cost.
We do this by removing the surrounding CNOTs and adding three increment/decrement operations to the circuit, as follows.
(Note that, to avoid cyclic dependencies, the increment and decrement constructions described in the next subsection will only use the same-register-size adder.)

First, because there are many target bits not reached by the time we have swept through the entire input register, carry signals aren't reaching the high bit of the target register.
We fix this by replacing the innermost CNOT, the carry-propagating CNOT, with a controlled increment.
Second and third, because we're using the MSB of the input register as the carry signal, it causes an increment at the LSB instead of at the correct position.
We undo the LSB increment with a controlled decrement, and use a controlled increment to apply the MSB's effect where it should have actually gone.
See the circuit diagram in \autoref{fig:inline-adder-into-large}.

@@

In the case of addition, we have the property $\lnot (\lnot x + K) = \lnot (-x - 1 + K) = -(-x - 1 + K) - 1 = x-K$.
So a valid inverting operation $H$ for addition is a NOT gate applied to every target wire; $H=X^{\otimes n}$.
If we then consider $G=\text{S}_{\text{-} K}$, where $S$ is an offset-by-subscript operation, we find that $[G, H] = [\text{S}_{-K}, X^{\otimes n}] = \text{S}_{2K}$.
We can control this constructed $\text{S}_{2K}$ operation by adding controls only to the multi-not operation $H$, which is cheap to do.
To add $K$ instead of $2K$, we temporarily prepend a dirty LSB onto the target register.
We show this construction in \autoref{fig:controlled-addition}.

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/offset.png}
  \caption{
      Offset circuit from \cite{haner2016}.
      $e$ is either 0 or 1.
      Uses $1$ dirty ancilla, $O(n \lg n)$ gates, and $O(n)$ depth (by overlapping the recursive cases).
  }
  \label{fig:offset}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{assets/inline-adder-into-large.png}
  \caption{
      Adder with target larger than source, using no ancillae.
      Uses $O(n)$ gates and depth.
      The increment and decrement gates all have at least one free wire to borrow as a dirty ancilla.
  }
  \label{fig:inline-adder-into-large}
\end{figure}
